K-means Clustering 86 Single Malt Scotch Whiskies
================================================
  
  The first time I had an Islay single malt, my mind was blown. In my first foray into the world of whiskies, I took the plunge into the smokiest, peatiest beast of them all -- Laphroig. That same night, dreams of owning a smoker were replaced by the desire to roam the landscape of smoky single malts. 

As an Islay fan, I wanted to investigate whether distilleries within a given region do in fact share taste characteristics. For this, I used a dataset profiling 86 distilleries based on 12 flavor categories.  

The data was obtained from [https://www.mathstat.strath.ac.uk/outreach/nessie/nessie_whisky.html](https://www.mathstat.strath.ac.uk/outreach/nessie/nessie_whisky.html).  


```{r, eval=T}
whiskies <- read.csv("data/whisky/whiskies.txt", row.names=1, stringsAsFactors = FALSE) 
```


I first went ahead and ensured that the dataset had no missing observations. I generated a subset of the data that included only the 12 flavor variables, rescaled for comparability using scale().


```{r, eval=T}
sum(is.na(whiskies)) # no missing observations
whiskies_k <- scale(whiskies[2:13]) # rescale selected vars for kmeans
```

K-means clustering assigns each observation membership to one of k clusters in such a way that minimizes the distance between each observation and it's cluster's mean. K-means clustering requires us to specify the number of clusters. Below, we iterate through kmeans() with clusters argument varying from 1 to maxCluster and plot the within groups sum of squares for each iteration.  


```{r, eval=T}
ssPlot <- function(data,maxCluster=9){
  # Initialize within sum of squares 
  SSw  <- (nrow(data)-1)*sum(apply(data,2,var))
  SSw <- vector()
  for (i in 2:maxCluster){
    SSw[i] <- sum(kmeans(data,centers=i)$withinss)
  }
  plot(1:maxCluster, SSw, type="b", 
       xlab="Number of Clusters",
       ylab="Within groups sum of squares") 
}
ssPlot(whiskies_k)
```

Naturally, the within groups sum of squares decreases as we increase the number of clusters. However, there is a trend of diminishing marginal returns as we increase the number of clusters. I select the number of clusters based on the point at which the marginal return of adding one more cluster is less than was the marginal return for adding the clusters prior to that.


```{r, eval=T}
fit <- kmeans(whiskies_k, 4) # 4 cluster solution 

# append cluster assignment
whiskies <- data.frame(whiskies, fit$cluster)
whiskies$fit.cluster <- as.factor(whiskies$fit.cluster)
```

Cluster centers can inform on how taste profiles differ between clusters.  

```{r, eval=T}
fit$centers 
```

Based on these centers, I anticipate that my love  for the full bodied, smoky and medicinal lies in cluster 4.

```{r, eval=T}
subset(whiskies, fit.cluster==4)
```

I identified the most representative whisky of each cluster by seeking out the observation closest to the center based on all 12 variables.


```{r, eval=T}
whiskies_r <- whiskies[c(2:13,17)] 
# extract just flavor variables & cluster
candidates <- by (whiskies_r[-13], whiskies_r[13], function(data){
  # we apply this function to observations for each level of fit.cluster
  dists <- sapply(data, function(x) (x-mean(x))^2) 
  # for each variable, calc each observation's deviation 
  # from average of the variable across observations 
  dists <- rowSums(dists) 
  # for each observation, sum the deviations across variables
  rownames(data)[dists==min(dists)] 
  # obtain the row number of the smallest sum
})

candidates <- as.numeric(unlist(candidates))

whiskies[candidates,]
```


The dataset contains coordinates that I used to investigate how flavor profiles differ geographically. The dataset's Latitude and Longitude variables are coordinates defined according to Great Britain's Ordnance Survey National Grid reference system. I converted the coordinates to standard latitude and longitude in order to plot them using ggmap.


```{r, eval=T}
library(maptools)
library(rgdal)

whiskies.coord <- data.frame(whiskies$Latitude,whiskies$Longitude)
coordinates(whiskies.coord)=~whiskies.Latitude+whiskies.Longitude

proj4string(whiskies.coord)=CRS("+init=epsg:27700") # Specify that our coords are in osgb grid coord

whiskies.coord <- spTransform(whiskies.coord,CRS("+init=epsg:4326")) # spTransform to convert osgb grid to lat/lon

whiskies <- cbind(whiskies,whiskies.coord)
```

Alternatively, the ggmap package ships with a geocode function which uses  Google Maps to determine the lat/lon based on a character string specifying the location.

```{r, eval=T}
library("ggmap")
```

```{r, eval=F}
whiskies <- cbind(whiskies, geocode(paste(whiskies$Location,"Scotland",sep=" ,")))
```


```{r, eval=T}

whiskyMap <- qmap(location = "Scotland", zoom = 6, legend = "topleft",
                  maptype = "terrain", color = "bw", darken=.5)


whiskyMap + geom_point(data=whiskies,
                       aes(x=whiskies.Latitude,
                           y=whiskies.Longitude,
                           colour=fit.cluster,
                           size=2))
```

I zoomed in and examine which Distilleries lie within the Islay region.

```{r}
whiskyMap <- qmap(location = "Islay", zoom = 10, legend = "topleft",
                  maptype = "terrain", color = "bw", darken=0.5)


whiskyMap + 
  geom_point()+
  geom_text(data=whiskies,
            aes(x=whiskies.Latitude,
                y=whiskies.Longitude,
                label=Distillery,
                color=fit.cluster,
                face="bold"))
```

The results indicate that there is a lot of variation in flavor profiles within the different scotch whisky regions. Note that initial cluster centers are chosen at random. In order to replicate the results, you will need to run the following code before your analysis. 

```{r, include=T, eval=F}
set.seed(1)
```

Further data analysis would be required to determine whether proximity to types of water sources or terrain types drive common flavor profiles. This could be done by obtaining shape files and adding them as an additional layer to the ggmap plot.

For me, I have identified my next to-try single malt. Talisker is still within the familiar realm of cluster 4 but a little more malty, fruity and spicy. Sounds like the perfect holiday mix.
